Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-16 paper/abstract.tex
small updates

The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
an age proxy, to reveal their rotational evolution and demonstrate that
kinematics could be a useful tool for calibrating gyrochronology.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
they are young, at old ages we find that late-G and early-K dwarfs rotate at
the {\it same rate} or faster than late-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that K dwarfs experience an epoch of stalled spin-down.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
K-dwarfs are likely to be synchronized binaries.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
an age proxy, to reveal their rotational evolution and demonstrate that
kinematics could be a useful tool for calibrating gyrochronology.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
they are young, at old ages we find that late-G and early-K dwarfs rotate at
the {\it same rate} or faster than late-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
K-dwarfs are likely to be synchronized binaries.

### 2020-03-16 paper/intro.tex
small updates

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} have revealed that rotational evolution is more
complicated than previously thought.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.
Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

>>>>

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} confirm that rotational evolution is a highly complex
process.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.
Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

### 2020-03-16 paper/results.tex
small updates

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' surface rotational
evolution of K dwarfs, like that observed in open clusters \citep{curtis2019}
and reproduced by models that vary angular momentum transport between stellar
core and envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, parker1970,
kawaler1988, charbonneau2010}.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the data presented in figure \ref{fig:vplot} however, at
intermediate ages, the rotation periods of K dwarfs appear {\it constant} with
mass, and at late ages rotation period {\it increases} with {\it increasing}
mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

>>>>

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' surface rotational
evolution of K dwarfs, like that observed in open clusters \citep{curtis2019}
and reproduced by models that vary angular momentum transport between stellar
core and envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010}.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.
